Species motif participation provides unique information about species risk of extinction

### 2023-05-19 manuscript/reorder_main_text.tex
Updates from Overleaf

    Motifs are unique patterns of $n$ interacting species that correspond to different sets of direct and indirect interactions, providing a holistic description of meso-scale network structure~\citep{Stouffer2007,Stouffer2012}.
    The vector of frequencies with which a species appears in each motif --its \emph{motif participation}-- describes the species' position in the meso-scale structure of a food web and, like degree and trophic level, is easily calculated once the structure of the network is known.
    Importantly, motif participation incorporates information on the positions of focal species' interaction partners, as well as potential indirect interactions between species.
    Two species might both have two prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists.
    Importantly, since specialists are usually more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than will a predator of generalists.
    This extra information on the focal species' interaction partners may thus provide extra insight when predicting extinction risk.

@@

    %There are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, 
    We define motif participation using the well-studied three-species motifs~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}. 
    Four of these motifs --the three-species chain, apparent competition, direct competition, and omnivory-- are especially interesting as they are by far the most common in empirical networks~\citep{Stouffer2007, giling2019plant} and more stable than other motifs (i.e., are more likely to retain all three species in the motif) when modelled in isolation~\citep{Borrelli2015a}.
    These four motifs also have clear ecological interpretations.
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The dynamics of three-species chains in isolation are well-understood from Lotka-Volterra models, and stable equilibria with all three species persisting can often be identified~\citep{Borrelli2015a,Hsu2015}.
    In general, food chains are more likely to be stable when predator-prey body-mass ratios are smaller~\citep{Jonsson1998}.
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers by depletion of the shared resource in competitive exclusion~\citep{Hsu2015}. 
    Despite the potential for exclusion where one predator is competitively superior, this motif is highly robust to changes in interaction strength and therefore has high quasi-sign stability~\citep{Borrelli2015a}.
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey indicates a generalist predator (which may be able persist following reduction or loss of one prey) but can capture indirect effects between prey that mae the predator more likely to go extinct~\citep{Melian2002}.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator. 
    Omnivory is less stable in isolation than the other three motifs~\citep{Borrelli2015a}, though there is a region of stable equilibrium for Lotka-Volterra model of omnivory~\citep{Hsu2015}.
    In whole food webs, frequent omnivory has been proposed to both increase and decrease overall stability in different conditions~\citep{McCann1997,Emmerson2004,Monteiro2016}.
    As most of the above studies either model motifs in isolation or examine overall effects on the whole food web, it remains unclear how a particular species' motif participation will affect its specific extinction risk.

>>>>

    Motifs are unique patterns of $n$ interacting species that correspond to different sets of direct and indirect interactions, providing a holistic description of meso-scale network structure~\citep{Stouffer2007,Stouffer2012}.
    Some motifs are over-represented in stable food webs, suggesting that participation in these motifs provides stability (refs).
    Indeed, dynamic simulations suggest that participation in these motifs can provide greater persistence after a disturbance~\citep{Cirtwill2022Oikos}.
    The vector of frequencies with which a species appears in each motif --its \emph{motif participation}-- describes the species' position in the meso-scale structure of a food web and, like degree and trophic level, is easily calculated once the structure of the network is known.
    Importantly, motif participation incorporates information on the positions of focal species' interaction partners, as well as potential indirect interactions between species.
    Two species might both have two prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists.
    Since specialists are usually more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than will a predator of generalists.
    This extra information on the focal species' interaction partners may thus provide extra insight when predicting extinction risk.

@@

    %There are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, 
    We define motif participation using the well-studied three-species motifs~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}. 
    Four of these motifs --the three-species chain, apparent competition, direct competition, and omnivory-- are especially interesting as they are by far the most common in empirical networks~\citep{Stouffer2007, giling2019plant} and more stable than other motifs (i.e., are more likely to retain all three species in the motif) when modelled in isolation~\citep{Borrelli2015a}.
    These four motifs also have clear ecological interpretations.
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The dynamics of three-species chains in isolation are well-understood from Lotka-Volterra models, and stable equilibria with all three species persisting can often be identified~\citep{Borrelli2015a,Hsu2015}.
    In general, food chains are more likely to be stable when predator-prey body-mass ratios are smaller~\citep{Jonsson1998}.
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers by depletion of the shared resource in competitive exclusion~\citep{Hsu2015}. 
    Despite the potential for exclusion where one predator is competitively superior, this motif is highly robust to changes in interaction strength and therefore has high quasi-sign stability~\citep{Borrelli2015a}.
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey. The motif indicates a generalist predator (which may be able persist following reduction or loss of one prey) but can capture indirect effects between prey that make the predator more likely to go extinct~\citep{Melian2002}.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator. 
    Omnivory is less stable in isolation than the other three motifs~\citep{Borrelli2015a}, though there is a region of stable equilibrium for Lotka-Volterra model of omnivory~\citep{Hsu2015}.
    In whole food webs, frequent omnivory has been proposed to both increase and decrease overall stability in different conditions~\citep{McCann1997,Emmerson2004,Monteiro2016}.
    As most of the above studies either model motifs in isolation or examine overall effects on the whole food web, it remains unclear how a particular species' motif participation will affect its specific extinction risk.
